---
title: "COVID-19's Impact on Motor Vehicle Collisions in NYC"
subtitle: "STAT 605 Final Presentation"
author: "Astha Singh, Franco Bosetti, Louis Clarke, Lindsey Russ"
format:
  revealjs: 
    # embed-resources: true
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    css: styles.css
    footer: "Group 1: Astha Singh, Franco Bosetti, Louis Clarke, Lindsey Russ"
server: shiny
---

```{r preprocessing, include=FALSE}
# Preprocessing 
library(RSQLite)
library(tidyr)
library(rcompanion)
library(ggplot2)
library(dplyr)
library(broom)
library(lubridate)
library(readr)
library(ggspatial)
library(sf)
library(prettymapr)
library(raster)
library(gridExtra)
library(ggeffects)
library(forcats)
library(stringr)
library(tidytext)
library(stringr)
data(stop_words)
library(sf)

dcon <- dbConnect(SQLite(), dbname = "~/Documents/STAT405_605/STAT605_Project/data/Collisions_DB.db")

# nyc preprocessing
dbExecute(dcon, "
CREATE VIEW IF NOT EXISTS crashes_enriched AS
WITH b AS (
  SELECT
    -- map raw columns to helper names
    substr(\"CRASH DATE\",7,4) || '-' || substr(\"CRASH DATE\",1,2) || '-' || substr(\"CRASH DATE\",4,2) AS crash_date_iso,
    \"CRASH TIME\"                                   AS crash_time_raw,
    \"BOROUGH\"                                      AS borough,
    \"ZIP CODE\"                                     AS zip_code,
    CAST(\"LATITUDE\"  AS REAL)                      AS latitude,
    CAST(\"LONGITUDE\" AS REAL)                      AS longitude,
    \"LOCATION\"                                     AS location,
    \"ON STREET NAME\"                               AS on_street_name,
    \"CROSS STREET NAME\"                            AS cross_street_name,
    \"OFF STREET NAME\"                              AS off_street_name,
    \"NUMBER OF PERSONS INJURED\"     AS number_of_persons_injured,
  \"NUMBER OF PERSONS KILLED\"      AS number_of_persons_killed,
  \"NUMBER OF PEDESTRIANS INJURED\" AS number_of_pedestrians_injured,
  \"NUMBER OF PEDESTRIANS KILLED\"  AS number_of_pedestrians_killed,
  \"NUMBER OF CYCLIST INJURED\"     AS number_of_cyclist_injured,
  \"NUMBER OF CYCLIST KILLED\"      AS number_of_cyclist_killed,
  \"NUMBER OF MOTORIST INJURED\"    AS number_of_motorist_injured,
  \"NUMBER OF MOTORIST KILLED\"     AS number_of_motorist_killed,
    \"CONTRIBUTING FACTOR VEHICLE 1\"                AS contributing_factor_vehicle_1,
    \"CONTRIBUTING FACTOR VEHICLE 2\"                AS contributing_factor_vehicle_2,
    \"CONTRIBUTING FACTOR VEHICLE 3\"                AS contributing_factor_vehicle_3,
    \"CONTRIBUTING FACTOR VEHICLE 4\"                AS contributing_factor_vehicle_4,
    \"CONTRIBUTING FACTOR VEHICLE 5\"                AS contributing_factor_vehicle_5,
    \"COLLISION_ID\"                                 AS collision_id,
    \"VEHICLE TYPE CODE 1\"                          AS vehicle_type_code_1,
    \"VEHICLE TYPE CODE 2\"                          AS vehicle_type_code_2,
    \"VEHICLE TYPE CODE 3\"                          AS vehicle_type_code_3,
    \"VEHICLE TYPE CODE 4\"                          AS vehicle_type_code_4,
    \"VEHICLE TYPE CODE 5\"                          AS vehicle_type_code_5
  FROM crashes
)
SELECT
  -- original columns with R-friendly names
  date(crash_date_iso)                               AS \"CRASH.DATE\",
  crash_time_raw                                     AS \"CRASH.TIME\",
  borough                                            AS \"BOROUGH\",
  zip_code                                           AS \"ZIP.CODE\",
  latitude                                           AS \"LATITUDE\",
  longitude                                          AS \"LONGITUDE\",
  location                                           AS \"LOCATION\",
  on_street_name                                     AS \"ON.STREET.NAME\",
  cross_street_name                                  AS \"CROSS.STREET.NAME\",
  off_street_name                                    AS \"OFF.STREET.NAME\",
  number_of_persons_injured                          AS \"NUMBER.OF.PERSONS.INJURED\",
  number_of_persons_killed                           AS \"NUMBER.OF.PERSONS.KILLED\",
  number_of_pedestrians_injured                      AS \"NUMBER.OF.PEDESTRIANS.INJURED\",
  number_of_pedestrians_killed                       AS \"NUMBER.OF.PEDESTRIANS.KILLED\",
  number_of_cyclist_injured                          AS \"NUMBER.OF.CYCLIST.INJURED\",
  number_of_cyclist_killed                           AS \"NUMBER.OF.CYCLIST.KILLED\",
  number_of_motorist_injured                         AS \"NUMBER.OF.MOTORIST.INJURED\",
  number_of_motorist_killed                          AS \"NUMBER.OF.MOTORIST.KILLED\",
  contributing_factor_vehicle_1                      AS \"CONTRIBUTING.FACTOR.VEHICLE.1\",
  contributing_factor_vehicle_2                      AS \"CONTRIBUTING.FACTOR.VEHICLE.2\",
  contributing_factor_vehicle_3                      AS \"CONTRIBUTING.FACTOR.VEHICLE.3\",
  contributing_factor_vehicle_4                      AS \"CONTRIBUTING.FACTOR.VEHICLE.4\",
  contributing_factor_vehicle_5                      AS \"CONTRIBUTING.FACTOR.VEHICLE.5\",
  collision_id                                       AS \"COLLISION_ID\",
  vehicle_type_code_1                                AS \"VEHICLE.TYPE.CODE.1\",
  vehicle_type_code_2                                AS \"VEHICLE.TYPE.CODE.2\",
  vehicle_type_code_3                                AS \"VEHICLE.TYPE.CODE.3\",
  vehicle_type_code_4                                AS \"VEHICLE.TYPE.CODE.4\",
  vehicle_type_code_5                                AS \"VEHICLE.TYPE.CODE.5\",
  -- derived fields
  strftime('%Y-%m', crash_date_iso)                  AS YEAR_MONTH,
  strftime('%Y',    crash_date_iso)                  AS YEAR,
  '(' || printf('%.4f', latitude) || ', ' ||
        printf('%.4f', longitude) || ')'             AS round_location,
  CASE
    WHEN instr(crash_time_raw, ':') > 0
      THEN CAST(substr(crash_time_raw, 1, instr(crash_time_raw, ':') - 1) AS INTEGER)
    ELSE NULL
  END                                                AS HOUR,
  CASE
    WHEN date(crash_date_iso) <  date('2020-03-01')                          THEN 'Pre-COVID'
    WHEN date(crash_date_iso) >= date('2020-03-01')
     AND date(crash_date_iso) <  date('2022-01-01')                          THEN 'During COVID'
    ELSE 'Post-COVID'
  END                                                AS covid_period,
  CASE
    WHEN date(crash_date_iso) >= date('2020-03-01')
     AND date(crash_date_iso) <  date('2022-01-01')                          THEN 1 ELSE 0
  END                                                AS is_covid,
  CASE
    WHEN date(crash_date_iso) >= date('2022-01-01')                          THEN 1 ELSE 0
  END                                                AS is_post_covid
FROM b
  WHERE date(crash_date_iso) >= date('2017-03-01')
;
")

dbExecute(dcon, "
CREATE VIEW IF NOT EXISTS vehicles_enriched AS
WITH v AS (
  SELECT
    -- original columns
    \"UNIQUE_ID\", \"COLLISION_ID\", \"CRASH_DATE\", \"CRASH_TIME\",
    \"VEHICLE_ID\", \"STATE_REGISTRATION\", \"VEHICLE_TYPE\", \"VEHICLE_MAKE\",
    \"VEHICLE_MODEL\", \"VEHICLE_YEAR\", \"TRAVEL_DIRECTION\", \"VEHICLE_OCCUPANTS\",
    \"DRIVER_SEX\", \"DRIVER_LICENSE_STATUS\", \"DRIVER_LICENSE_JURISDICTION\", \"PRE_CRASH\",
    \"POINT_OF_IMPACT\", \"VEHICLE_DAMAGE\", \"VEHICLE_DAMAGE_1\", \"VEHICLE_DAMAGE_2\",
    \"VEHICLE_DAMAGE_3\", \"PUBLIC_PROPERTY_DAMAGE\", \"PUBLIC_PROPERTY_DAMAGE_TYPE\",
    \"CONTRIBUTING_FACTOR_1\", \"CONTRIBUTING_FACTOR_2\",

    -- helpers
    substr(\"CRASH_DATE\",7,4) || '-' || substr(\"CRASH_DATE\",1,2) || '-' || substr(\"CRASH_DATE\",4,2) AS crash_date_iso,
    CAST(\"VEHICLE_YEAR\" AS INTEGER)      AS veh_year_int,
    CAST(\"VEHICLE_OCCUPANTS\" AS INTEGER) AS veh_occ_int
  FROM vehicles
)
SELECT
  v.*,
  CAST(strftime('%Y', crash_date_iso) AS INTEGER)                  AS year_int,
  CASE
    WHEN date(crash_date_iso) <  date('2020-03-01')                THEN 'Pre-COVID'
    WHEN date(crash_date_iso) >= date('2020-03-01')
     AND date(crash_date_iso) <  date('2022-01-01')                THEN 'During COVID'
    ELSE 'Post-COVID'
  END                                                              AS covid_period
FROM v;
")

dbExecute(dcon, "
CREATE VIEW IF NOT EXISTS vehicle_agg AS
SELECT
  COLLISION_ID,
  COUNT(*)                                                     AS VEHICLE_COUNT,
  AVG(CASE
        WHEN veh_occ_int BETWEEN 1 AND 60                      THEN veh_occ_int
      END)                                                     AS AV_PASSENGERS,
  AVG(CASE
        WHEN veh_year_int BETWEEN 1920 AND 2025
         AND veh_year_int <= year_int                          THEN veh_year_int
      END)                                                     AS AV_VEH_YEAR
FROM vehicles_enriched
GROUP BY COLLISION_ID;
")

nyc <- dbGetQuery(dcon, "
SELECT
  ce.*,                            
  va.VEHICLE_COUNT,
  va.AV_PASSENGERS,
  va.AV_VEH_YEAR,
  CAST(ce.YEAR AS INTEGER) - va.AV_VEH_YEAR AS AV_VEH_AGE
FROM crashes_enriched AS ce
LEFT JOIN vehicle_agg   AS va
  ON ce.COLLISION_ID = va.COLLISION_ID;
")

nyc$covid_period <- factor(nyc$covid_period, 
                           levels = c("Pre-COVID","During COVID","Post-COVID"))

nyc$BOROUGH <- factor(nyc$BOROUGH)

nyc <- nyc %>%
mutate(across(
    c(NUMBER.OF.PERSONS.INJURED,
      NUMBER.OF.PERSONS.KILLED,
      NUMBER.OF.PEDESTRIANS.INJURED,
      NUMBER.OF.PEDESTRIANS.KILLED,
      NUMBER.OF.CYCLIST.INJURED,
      NUMBER.OF.CYCLIST.KILLED,
      NUMBER.OF.MOTORIST.INJURED,
      NUMBER.OF.MOTORIST.KILLED),
    as.integer
  ))

# vzv speed limit preprocessing
# Format the sp_limits latitude and longitude
# Pre-process the speed limit data
dbExecute(dcon, "
  DROP VIEW IF EXISTS v_speed_limits_parsed;
")
dbExecute(dcon, "
  DROP VIEW IF EXISTS v_sp_lim_proc;
")
dbExecute(dcon, "
  DROP VIEW IF EXISTS v_sp_limits_filt;
")

dbExecute(dcon, "
CREATE TEMP VIEW v_speed_limits_parsed AS
WITH base AS (
  SELECT
    rowid AS rid,
    the_geom,
    street,
    postvz_sl,
    postvz_sg,
    Shape_Leng,
    CASE
      WHEN the_geom LIKE 'POINT (%'
        THEN TRIM(REPLACE(REPLACE(the_geom, 'POINT (', ''), ')', ''))
      WHEN the_geom LIKE 'LINESTRING (%'
        THEN TRIM(REPLACE(REPLACE(the_geom, 'LINESTRING (', ''), ')', ''))
      WHEN the_geom LIKE 'MULTILINESTRING ((%'
        THEN TRIM(REPLACE(REPLACE(the_geom, 'MULTILINESTRING ((', ''), '))', ''))
      ELSE NULL
    END AS raw_coords
  FROM speed_limits
),
first_pair AS (
  SELECT
    rid, the_geom, street, postvz_sl, postvz_sg, Shape_Leng,
    CASE
      WHEN the_geom LIKE 'POINT (%'
        THEN raw_coords
      ELSE
        SUBSTR(raw_coords, 1,
          CASE WHEN INSTR(raw_coords, ',') > 0
               THEN INSTR(raw_coords, ',') - 1
               ELSE LENGTH(raw_coords)
          END)
    END AS inner_xy
  FROM base
),
split_xy AS (
  SELECT
    rid, the_geom, street, postvz_sl, postvz_sg, Shape_Leng, inner_xy,
    CAST(SUBSTR(inner_xy, 1, INSTR(inner_xy, ' ') - 1) AS REAL) AS longitude,
    CAST(SUBSTR(inner_xy, INSTR(inner_xy, ' ') + 1) AS REAL)   AS latitude
  FROM first_pair
)
SELECT
  rid, the_geom, street, postvz_sl, postvz_sg, Shape_Leng, latitude, longitude
FROM split_xy;
")

dbExecute(dcon, "
CREATE TEMP VIEW v_sp_lim_proc AS
SELECT
  rid,
  the_geom,
  street,
  CAST(postvz_sl AS REAL) AS SPEED_LIMIT,
  postvz_sg,
  Shape_Leng,
  latitude,
  longitude,
  '(' || printf('%.4f', latitude) || ', ' || printf('%.4f', longitude) || ')' AS ROUND_LOCATION
FROM v_speed_limits_parsed;
")


dbExecute(dcon, "
CREATE TEMP VIEW v_sp_limits_filt AS
WITH ranked AS (
  SELECT
    *,
    ROW_NUMBER() OVER (
      PARTITION BY ROUND_LOCATION
      ORDER BY SPEED_LIMIT DESC, rid ASC
    ) AS rn
  FROM v_sp_lim_proc
)
SELECT *
FROM ranked
WHERE rn = 1;
")

sp_limits_filt <- dbGetQuery(dcon, "SELECT the_geom, street, SPEED_LIMIT AS \"SPEED.LIMIT\", postvz_sg, Shape_Leng, latitude, longitude, ROUND_LOCATION AS \"ROUND.LOCATION\" FROM v_sp_limits_filt;")

combined <- # add speed limit data to nyc dataset
  mutate(nyc, ROUND.LOCATION = paste("(", round(LATITUDE, digits = 4), ", ", 
                                     round(LONGITUDE, digits = 4), ")", 
                                     sep = "")) %>%
  filter(!is.na(LATITUDE), !is.na(LONGITUDE)) %>%
  inner_join(dplyr::select(sp_limits_filt, SPEED.LIMIT, ROUND.LOCATION), 
             by="ROUND.LOCATION")

# twitter preprocessing 2
# transit vs vehicle tweets
dbExecute(dcon, "DROP VIEW IF EXISTS v_twitter_flags;")
dbExecute(dcon, "DROP VIEW IF EXISTS v_twitter_vehicle;")
dbExecute(dcon, "DROP VIEW IF EXISTS v_twitter_nyc;")

dbExecute(dcon, "
CREATE TEMP VIEW v_twitter_nyc AS
WITH base AS (
  SELECT
    class,
    tweet_id,
    tweet_text,
    LOWER(tweet_text) AS txt
  FROM tweets
  WHERE class IN ('1','2')
),
tagged AS (
  SELECT
    class,
    tweet_id,
    tweet_text,
    CASE
      WHEN txt LIKE '%manhattan%' THEN 'Manhattan'
      WHEN txt LIKE '%brooklyn%' THEN 'Brooklyn'
      WHEN txt LIKE '%queens%' THEN 'Queens'
      WHEN txt LIKE '%bronx%' THEN 'Bronx'
      WHEN txt LIKE '%staten island%' THEN 'Staten Island'
      WHEN txt LIKE '%nyc%'
        OR txt LIKE '%new york city%'
        OR txt LIKE '%new york%' THEN 'NYC (unspecified)'
      ELSE NULL
    END AS location
  FROM base
)
SELECT * FROM tagged WHERE location IS NOT NULL;
")

dbExecute(dcon, "
CREATE TEMP VIEW v_twitter_flags AS
SELECT
  class,
  tweet_id,
  tweet_text,
  location,

  CASE
    WHEN LOWER(tweet_text) LIKE '%subway%'
      OR LOWER(tweet_text) LIKE '%train%'
      OR LOWER(tweet_text) LIKE '%#%line%'
      OR LOWER(tweet_text) LIKE '%station%'
      OR LOWER(tweet_text) LIKE '%mta%'
      OR LOWER(tweet_text) LIKE '%bus%'
    THEN TRUE ELSE FALSE
  END AS is_transit,

  CASE
    WHEN LOWER(tweet_text) LIKE '%car%'
      OR LOWER(tweet_text) LIKE '%vehicle%'
      OR LOWER(tweet_text) LIKE '%crash%'
      OR LOWER(tweet_text) LIKE '%accident%'
      OR LOWER(tweet_text) LIKE '%collision%'
      OR LOWER(tweet_text) LIKE '%highway%'
      OR LOWER(tweet_text) LIKE '%road%'
      OR LOWER(tweet_text) LIKE '%traffic%'
    THEN TRUE ELSE FALSE
  END AS is_vehicle
FROM v_twitter_nyc;
")

dbExecute(dcon, "
CREATE TEMP VIEW v_twitter_vehicle AS
SELECT
  *,
  CASE
    WHEN (LOWER(tweet_text) LIKE '%accident%' OR LOWER(tweet_text) LIKE '%crash%'
          OR LOWER(tweet_text) LIKE '%collision%' OR LOWER(tweet_text) LIKE '%wreck%')
         AND LOWER(tweet_text) NOT LIKE '%subway%'
         AND LOWER(tweet_text) NOT LIKE '%train%'
         AND LOWER(tweet_text) NOT LIKE '%#%line%'
         AND LOWER(tweet_text) NOT LIKE '%station%'
    THEN TRUE ELSE FALSE
  END AS is_real_crash,
  CASE
    WHEN (LOWER(tweet_text) LIKE '%highway%' OR LOWER(tweet_text) LIKE '%parkway%'
          OR LOWER(tweet_text) LIKE '%expressway%' OR LOWER(tweet_text) LIKE '%west side%'
          OR LOWER(tweet_text) LIKE '%fdr%')
         AND LOWER(tweet_text) NOT LIKE '%station%'
    THEN TRUE ELSE FALSE
  END AS is_road_incident
FROM v_twitter_flags
WHERE
  ((LOWER(tweet_text) LIKE '%accident%' OR LOWER(tweet_text) LIKE '%crash%'
     OR LOWER(tweet_text) LIKE '%collision%' OR LOWER(tweet_text) LIKE '%wreck%')
     AND LOWER(tweet_text) NOT LIKE '%subway%'
     AND LOWER(tweet_text) NOT LIKE '%train%'
     AND LOWER(tweet_text) NOT LIKE '%#%line%'
     AND LOWER(tweet_text) NOT LIKE '%station%')
  OR
  ((LOWER(tweet_text) LIKE '%highway%' OR LOWER(tweet_text) LIKE '%parkway%'
     OR LOWER(tweet_text) LIKE '%expressway%' OR LOWER(tweet_text) LIKE '%west side%'
     OR LOWER(tweet_text) LIKE '%fdr%')
     AND LOWER(tweet_text) NOT LIKE '%station%');
")


twitter_vehicle <- dbGetQuery(dcon, "SELECT * FROM v_twitter_vehicle") %>%
  mutate(
    is_transit = as.logical(is_transit),
    is_vehicle = as.logical(is_vehicle),
    is_real_crash = as.logical(is_real_crash),
    is_road_incident = as.logical(is_road_incident)
  )

dbDisconnect(dcon)
```


# Introduction {background="#6B6B6B"}

## Project Overview {.smaller}

**Research Question:** How did COVID-19 affect motor vehicle collision patterns in New York City?

::: incremental
- **Dataset:** 2.22M police-reported crashes (2012-2025)
- **Time Periods:** 
  - Pre-COVID (before March 2020)
  - During COVID (March 2020 - Dec 2021)
  - Post-COVID (2022-present)
- **Key Questions:**
  1. Did collision rates return to pre-pandemic levels?
  2. Did crashes become more deadly?
  3. What changed in driver behavior?
:::

## Data Sources {.smaller}

::: columns
::: {.column width="50%"}
**Datasets:**

- **Primary:** NYC Motor Vehicle Collisions (2.22M records)
- **Secondary:** Vehicle-Level Details (4.45M records)
- **Tertiary:** NYC Speed Limits (141K segments)
- **Quaternary:** Twitter Traffic Reports (798 NYC tweets)
:::

::: {.column width="50%"}
**Key Variables:**

- Crash date/time & location
- Casualties by road user type
- Contributing factors (up to 5)
- Vehicle characteristics
- Borough information
:::
:::

# Did Collision Rates Change? {background="#43464B"}

## Monthly Collision Trends {.smaller}

**Number of Monthly Vehicle Collisions in NYC:**
```{r plot1, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Motor Vehicle Collisions over Time
par(mfrow = c(1, 2))

monthly_counts <- table(nyc$YEAR_MONTH)
months <- names(monthly_counts)
colspermonth <- as.numeric(monthly_counts)
plot_months <- as.Date(paste0(months, "-01"))

par(mar = c(5, 5, 4, 2) + 0.1) 

plot(plot_months, colspermonth,
     cex.main = 0.9,
     type="n",
     pch=20,
     xlab="Month",
     ylab="Number of Collisions",
     xaxt="n",
     yaxt="n",
     xaxs="i",
     yaxs="i",
     ylim=c(0,max(colspermonth)+1000),
     mgp = c(3, 1, 0))

rect(xleft = as.Date("2017-01-01"), 
     ybottom = 0, 
     xright = as.Date("2020-01-01"), 
     ytop = max(colspermonth)+1000,
     col = adjustcolor("lightblue", alpha.f = 0.3), 
     border = NA)

rect(xleft = as.Date("2022-01-01"), 
     ybottom = 0, 
     xright = as.Date("2025-01-01"), 
     ytop = max(colspermonth)+1000,
     col = adjustcolor("lightgreen", alpha.f = 0.3), 
     border = NA)

points(plot_months, colspermonth, pch=20)
lines(plot_months, colspermonth)

xticks <- seq(min(plot_months), max(plot_months)+365, by = "6 months")
axis.Date(1, at = xticks, 
          format = "%b\n%Y", cex.axis = 0.6)
axis(2, at = seq(0, max(colspermonth)+1000, by = 1000), cex.axis = 0.7, las=1) 
abline(v = xticks, col = "gray", lty = "dotted")
abline(v=as.Date("2020-03-17"), col="red", lwd=1.5, lty=5)
text(x=as.Date("2020-03-17"), y=max(colspermonth)*0.9, 
     labels="Beginning of \nCOVID lockdown", pos=4, col="red", cex=0.8)


borough <- levels(nyc$BOROUGH)
borough <- borough[borough != ""]
colors = c("red", "blue", "darkgreen", "purple", "orange")
plot_months <- list()
colspermonth <- list()
for (b in borough){
  monthly_counts <- table(nyc$YEAR_MONTH[nyc$BOROUGH==b])
  months <- names(monthly_counts)
  colspermonth[[b]] <- as.numeric(monthly_counts)
  plot_months[[b]] <- as.Date(paste0(months, "-01"))
}

max_col <- max(sapply(colspermonth, max))
min_dat <- as.Date(min(sapply(plot_months,min)))
max_dat <- as.Date(max(sapply(plot_months,max)))

plot(plot_months[[1]], colspermonth[[1]],
     type="l",
     pch=20,
     col= colors[1],
     xlab="Month",
     ylab="Number of Collisions",
     xaxt="n",
     yaxt="n",
     xaxs="i",
     yaxs="i",
     ylim=c(0,max_col+500))

for (i in 2:(length(borough))){
  lines(plot_months[[i]], colspermonth[[i]], type="l", pch=20, col=colors[i])
}

legend("topright", 
       legend = borough,
       col = colors,
       lty = 1,
       cex = 0.7,
       inset = c(0.02, 0.02),
       bg = "white",
       box.lty = 0)

xticks <- seq(min_dat, max_dat+365, by = "6 months")
axis.Date(1, at = xticks, 
          format = "%b\n%Y", cex.axis = 0.6)
axis(2, at = seq(0, max_col+1000, by = 1000), cex.axis = 0.7, las=1) 
abline(v = xticks, col = "gray", lty = "dotted")
abline(v=as.Date("2020-03-17"), col="black", lwd=1.5, lty=5)
text(x=as.Date("2020-03-17"), y=max_col*0.7, 
     labels="Beginning of COVID lockdown", pos=4, col="black", cex=0.6)

par(mfrow = c(1, 1))
```


## Borough-Level Patterns {.smaller}

```{r plot2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
dat <- nyc %>%
  filter(BOROUGH != "") %>%
  group_by(BOROUGH) %>%
  count(YEAR) %>%
  ungroup() %>%
  mutate(period = ifelse(YEAR <= 2019, "2019 and before", "2020 and after"))

ggplot(dat, aes(x = YEAR, y = n, color = BOROUGH)) +
  geom_point() +
  geom_smooth(aes(linetype = period, group = interaction(BOROUGH, period)),
              method = "lm",
              se = FALSE) +
  labs(title = "Piecewise Linear Trends in Total Crashes per Year by Borough",
       x = "Year",
       y = "Number of Crashes",
       linetype = "Time Period")
```

::: {.fragment}
- **Brooklyn & Queens & Bronx:** Increasing pre-COVID, Decreasing post-COVID
- **Manhattan:** Decreasing pre-COVID, more static post-COVID
:::

# Did Crashes Become More Severe? {background="#43464B"}

## Severity  {.smaller}

```{r plot3, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Crash Severity by COVID Period
crash_severity_period <- nyc %>%
  group_by(covid_period) %>%
  summarize(
    fatal_rate = sum(NUMBER.OF.PERSONS.KILLED, na.rm = TRUE) / n() * 100000,
    injury_rate = sum(NUMBER.OF.PERSONS.INJURED, na.rm = TRUE) / n() * 100000
  ) %>%
  pivot_longer(
    cols = c(fatal_rate, injury_rate),
    names_to = "type",
    values_to = "rate"
  )

ggplot(crash_severity_period, aes(x = covid_period, y = rate, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("fatal_rate" = "firebrick", "injury_rate" = "steelblue"),
    labels = c("Fatality Rate", "Injury Rate")
  ) +
  facet_wrap(
    ~type,
    scales = "free_y",
    labeller = as_labeller(c(
      fatal_rate = "Fatality Rate",
      injury_rate = "Injury Rate"
    ))
  ) +
  labs(
    title = "Crash Severity by COVID Period",
    x = "COVID Period",
    y = "Rate per 100,000 Crashes",
    fill = "Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

::: {.fragment}
**Key Finding:** Crash severity increased during and after COVID‑19
:::

## COVID's Impact on Severity {.smaller}
```{r plot4, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
nyc_small <- nyc %>%
  transmute(
    covid_period = covid_period,
    is_fatal = ifelse(NUMBER.OF.PERSONS.KILLED > 0, 1, 0),
    borough = BOROUGH,
    crash_hour = hour(hm(CRASH.TIME)),
    contributing_factor_grouped = factor(case_when(
    grepl("ALCOHOL|DRUGS|ILLNES|FATIGUE|ASLEEP|CONSCIOUSNESS|MEDICATION|DISABILITY", 
          CONTRIBUTING.FACTOR.VEHICLE.1, ignore.case = TRUE) ~ "Impairment",
    grepl("SPEED|AGGRESSIVE|RAGE", 
          CONTRIBUTING.FACTOR.VEHICLE.1, ignore.case = TRUE) ~ "Speeding/Aggression",
    grepl("DISTRACTION|INATTENTION|PHONE|TEXTING|ELECTRONIC|NAVIGATION|EATING|HEADPHONES", 
          CONTRIBUTING.FACTOR.VEHICLE.1, ignore.case = TRUE) ~ "Distraction",
    grepl("YIELD|TRAFFIC CONTROL|KEEP RIGHT", 
          CONTRIBUTING.FACTOR.VEHICLE.1, ignore.case = TRUE) ~ "Traffic Violation",
    grepl("FOLLOWING|PASSING|LANE|TURNING|BACKING|UNINVOLVED|EXPERIENCE|PEDESTRIAN ERROR", 
          CONTRIBUTING.FACTOR.VEHICLE.1, ignore.case = TRUE) ~ "Driver Error",
    grepl("PAVEMENT|GLARE|VIEW|OBSTRUCT|ANIMAL|DEFECTIVE|FAILURE|TINTED|WINDSHIELD|OVERSIZED|VANDALISM", 
          CONTRIBUTING.FACTOR.VEHICLE.1, ignore.case = TRUE) ~ "External Vehicle Defect",
    TRUE ~ "Other/Unknown"
    )),
    vehicle_cat = factor(case_when(
      grepl("MOTORCYCLE|MOTORBIKE", VEHICLE.TYPE.CODE.1, ignore.case = TRUE) ~ "Motorcycle",
      grepl("TRUCK|BUS", VEHICLE.TYPE.CODE.1, ignore.case = TRUE) ~ "Truck/Bus",
      grepl("BICYCLE|BIKE|SCOOTER", VEHICLE.TYPE.CODE.1, ignore.case = TRUE) ~ "Bike/Scooter",
      TRUE ~ "Car/SUV"
    ))
  ) %>%
  filter(borough != "") %>%
  # Add Time of Day logic
  mutate(tod = ifelse(crash_hour >= 6 & crash_hour < 18, "day", "night"))

# 1. Set Factor Baseline (The "Normal" Crash)
nyc_small$contributing_factor_grouped <- relevel(nyc_small$contributing_factor_grouped, ref = "Other/Unknown")

# 2. Set Time Baseline (The "Normal" Time)
nyc_small$covid_period <- relevel(nyc_small$covid_period, ref = "Pre-COVID")

# Model 2: COVID period controlling for key factors
covid_adjusted_model <- glm(is_fatal ~ covid_period + borough + vehicle_cat + tod,
                           data = nyc_small, 
                           family = binomial)

covid_results <- tidy(covid_adjusted_model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(grepl("covid_period", term)) %>%
  mutate(period = case_when(
    grepl("During COVID", term) ~ "During COVID",
    grepl("Post-COVID", term) ~ "Post-COVID",
    TRUE ~ "Reference"
  )) %>%
  filter(period != "Reference")

ggplot(covid_results, aes(x = period, y = estimate)) +
  geom_point(size = 5, color = "#D7191C") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, size = 1.5) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", size = 1) +
  geom_text(aes(label = sprintf("%.2fx", estimate)),
            hjust = -0.5, vjust = 0.3, size = 6, fontface = "bold") +
  labs(title = "Fatal Crash Risk Rose during COVID",
       subtitle = "Adjusted for vehicle type, time, and location",
       x = "COVID Period", y = "Odds Ratio (vs. Pre-COVID)") +
  ylim(1, 2.5) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

::: {.fragment}
**Key Finding:** Even controlling for other factors, crashes became 67-78% more likely to be fatal
:::

## Crash Timing Patterns {.smaller}

```{r plot5, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Hourly Crash Distribution by COVID Period
hourly_proportions <- nyc %>%
  group_by(covid_period) %>%
  count(HOUR) %>%
  mutate(prop = n / sum(n))

ggplot(hourly_proportions, aes(x = HOUR, y = prop, fill = covid_period)) +
  geom_col(position = "dodge") +
  scale_x_continuous(
    breaks = seq(0, 23, by = 3),
    labels = c("12AM", "3AM", "6AM", "9AM", "12PM", "3PM", "6PM", "9PM")
  ) +
  labs(
    title = "Proportional Crash Distribution by Hour Across COVID Periods",
    y = "Proportion of Total Crashes in Period",
    x = "Hour of Day"
  ) +
  theme_minimal()

```

::: {.fragment}
Rush‑hour peaks dipped slightly during COVID but largely returned afterward
:::


# What Caused These Changes? {background="#43464B"}

## Contributing Factors {.smaller}

**How Factors Changed:**
```{r plot6, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
all_factors <- toupper(c(
  nyc$CONTRIBUTING.FACTOR.VEHICLE.1,
  nyc$CONTRIBUTING.FACTOR.VEHICLE.2,
  nyc$CONTRIBUTING.FACTOR.VEHICLE.3,
  nyc$CONTRIBUTING.FACTOR.VEHICLE.4,
  nyc$CONTRIBUTING.FACTOR.VEHICLE.5
))

factor_counts <- table(all_factors)
# removing all unspecified 
factor_counts <- factor_counts[!names(factor_counts) %in% 
                                c("", "UNSPECIFIED", "1", "80")]
top_5_factors <- names(sort(factor_counts, decreasing = TRUE)[1:5])

# Now create data with standardized factors
factor_data <- nyc %>%
  mutate(across(starts_with("CONTRIBUTING.FACTOR.VEHICLE."), toupper)) %>%
  pivot_longer(
    cols = starts_with("CONTRIBUTING.FACTOR.VEHICLE."),
    names_to = "vehicle_num",
    values_to = "factor"
  ) %>%
  filter(
    factor %in% top_5_factors,
    !is.na(covid_period)
  ) %>%
  count(covid_period, factor) %>%
  group_by(covid_period) %>%
  mutate(
    total = sum(n),
    pct = n / total * 100
  ) %>%
  ungroup()

ggplot(factor_data, aes(x = reorder(factor, -pct), y = pct, fill = covid_period)) +
  geom_col(position = position_dodge(preserve = "single"), color = "black", linewidth = 0.5) +
  scale_fill_manual(
    values = c("Pre-COVID" = "#F8766D",
               "During COVID" = "darkseagreen3",
               "Post-COVID" = "#619CFF"),
    name = "Period: "
  ) +
  labs(
    title = "Top Causes by COVID Period",
    x = NULL,
    y = "% of Total Crashes"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1, size = 5, color = "black"),
    axis.text.y = element_text(size = 8, color = "black"),
    axis.title.y = element_text(color = "black"),
    plot.title = element_text(face = "bold", size = 11, color = "black"),
    legend.position = "top",
    legend.text = element_text(color = "black"),
    legend.title = element_text(color = "black")
  )
```

::: {.fragment}
**Key Finding:** Driver inattention/distraction dominates, but its relative importance increased during COVID
:::

## Text Mining Insights {.smaller}

```{r plot7}

# Twitter words
twitter_words <- twitter_vehicle %>%
  unnest_tokens(word, tweet_text) %>%
  anti_join(stop_words, by = "word") %>%
  filter(!str_detect(word, "^\\d+$"),
         !word %in% c("nyc","new","york","updated","incident","cleared",
                      "directions","construction","highway","exit","amp",
                      "https","t.co")) %>%
  count(word, sort = TRUE) %>%
  mutate(percentage = n / sum(n) * 100, source = "Twitter") %>%
  head(12)

# NYC factors
factor_words <- nyc %>%
  pivot_longer(cols = starts_with("CONTRIBUTING.FACTOR.VEHICLE."),
               values_to = "factor", values_drop_na = TRUE) %>%
  filter(factor != "", factor != "Unspecified") %>%
  unnest_tokens(word, factor) %>%
  anti_join(stop_words, by = "word") %>%
  filter(!word %in% c("driver","vehicle","inattention","error","unspecified")) %>%
  count(word, sort = TRUE) %>%
  mutate(percentage = n / sum(n) * 100, source = "Motor Vehicle Collisions") %>%
  head(12)

bind_rows(twitter_words, factor_words) %>%
  ggplot(aes(x = percentage, y = reorder_within(word, percentage, source),
             fill = source)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~source, scales = "free") +
  scale_y_reordered() +
  scale_fill_manual(values = c("Twitter" = "lightblue",
                               "Motor Vehicle Collisions" = "lightgreen")) +
  labs(title = "Text Mining: Twitter vs. Motor Vehicle Collisions",
       x = "Percentage of Terms (%)", y = NULL) +
  theme_minimal(base_size = 14)
```


# Statistical Modeling {background="#43464B"}

## Baseline Risk Factors {.smaller}

```{r plot8, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
model_interaction <- glm(is_fatal ~ borough + vehicle_cat + tod + 
                        (contributing_factor_grouped * covid_period),
                        data = filter(nyc_small, 
                                      covid_period != "During COVID"), 
                        family = binomial)

# Pre-COVID Model
model_pre <- glm(is_fatal ~ borough + vehicle_cat + tod + contributing_factor_grouped, 
                 data = filter(nyc_small, covid_period == "Pre-COVID"), 
                 family = binomial)

# Post-COVID Model
model_post <- glm(is_fatal ~ borough + vehicle_cat + tod + contributing_factor_grouped, 
                  data = filter(nyc_small, covid_period == "Post-COVID"), 
                  family = binomial)

#Control
controls_results <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(!grepl(":", term)) %>%             # Remove Interaction terms
  filter(term != "(Intercept)") %>%         # Remove Intercept
  filter(!grepl("covid_period", term)) %>%  # Remove the general Time Period term
  filter(!grepl("contributing_factor", term)) %>% # Remove Factors (saved for the other plot)
  mutate(
    term_clean = case_when(
      grepl("borough", term) ~ gsub("borough", "Borough: ", term),
      grepl("vehicle_cat", term) ~ gsub("vehicle_cat", "Vehicle: ", term),
      grepl("tod", term) ~ gsub("tod", "Time: ", term),
      TRUE ~ term
    )
  )

ggplot(controls_results, aes(x = estimate, y = fct_reorder(term_clean, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "gray40") +
  geom_point(size = 3, color = "#2c3e50") +
  scale_x_log10() +
  labs(
    title = "Baseline Control Factors: What increases Fatal Risk?",
    x = "Odds Ratio (Log Scale)",
    y = NULL,
    caption = "NOTE: These effects are held constant in the model."
  ) +
  theme_minimal(base_size = 12)

```


::: {.fragment}
- **Motorcycles:** 3x+ more likely to be fatal
- **Nighttime:** ~2x higher odds than daytime
- **Manhattan:** Lower risk due to congestion
:::

## How COVID Changed Risk {.smaller}

```{r plot9, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
interaction_results <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(grepl(":", term)) %>%
  mutate(
    term_clean = gsub("contributing_factor_grouped", "", term),
    term_clean = gsub(":covid_periodPost-COVID", "", term_clean),
    is_significant = ifelse(p.value < 0.05, "Significant Shift", "No Significant Change")
  )

tidy_pre <- tidy(model_pre, exponentiate = TRUE, conf.int = TRUE) %>% 
  mutate(period = "Pre-COVID")
tidy_post <- tidy(model_post, exponentiate = TRUE, conf.int = TRUE) %>% 
  mutate(period = "Post-COVID")

plot_data <- bind_rows(tidy_pre, tidy_post) %>%
  filter(grepl("contributing_factor", term)) %>%
  # Clean up the names for the chart
  mutate(term = gsub("contributing_factor_grouped", "", term))

p1 <- ggplot(plot_data, aes(y = term, x = estimate)) +
  # The line connecting the dots
  geom_line(aes(group = term), color = "grey50", size = 1) + 
  # The dots themselves
  geom_point(aes(color = period), size = 4) + 
  # Add a vertical line at 1 (The Baseline/Unspecified Risk)
  geom_vline(xintercept = 1, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("Pre-COVID" = "#1f77b4", "Post-COVID" = "#d62728")) +
  labs(
    title = "How Did Fatal Crash Risks Shift?",
    x = "Odds Ratio (Log Scale)",
    y = NULL,
    color = "Period"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 14)

p2 <- ggplot(interaction_results, aes(x = estimate, y = fct_reorder(term_clean, estimate), color = is_significant)) +
  
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.7) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, linewidth = 0.8) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Significant Shift" = "#D55E00", "No Significant Change" = "gray70")) +
  scale_x_log10() +
  labs(
    title = "Did the Risk Change Significantly Post-COVID?",
    x = "Change in Odds Ratio (Log Scale)\n(Right = Became Riskier | Left = Became Safer)",
    y = "Risk Factor",
    color = "Statistical Significance"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

grid.arrange(p1, p2, ncol = 2)
```

# Killer Plot {background="#43464B"}

## Interactive Crash Density Map {.smaller}
```{r plot10}
#| panel: sidebar
sliderInput('n', 'Number of points', 50, min = 10, max = 100)
```
```{r}
#| panel: fill
plotOutput('plot')
```
```{r}
#| context: server
output$plot <- renderPlot({
  plot(1:input$n, rnorm(input$n))
})
```

# Conclusion {background="#43464B"}

## Key Findings {.smaller}

::: incremental
1. Collision trends

   - Rates stayed lower than before COVID and never fully returned
   - Each borough showed a distinct pattern; Brooklyn and Queens fell hardest, Manhattan recovered faster

2. Severity trends
   - Fatality rate per crash rose sharply during and after COVID
   - Injuries per crash fell, indicating fewer but more lethal crashes

3. Driver‑behavior
   - Speeding/Aggression and Impairment remained high‑risk with unchanged odds
   - Distraction and Traffic‑Violation odds fell significantly
::: 


# Thank You! {background="#43464B"}

